sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-02 doc/paper.Rmd
X_3,t
Clean up;edit

<!-- # The sdmTMB package implements spatiotemporal predictive-process GLMMs (Generalized Linear Mixed Effect Models) using TMB, INLA, and Gaussian Markov random fields. One common application is for spatial or spatiotemporal species distribution models (SDMs). -->

@@

*[quick draft outline of intro; may be too long]*

@@

Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA.
Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through X case studies, and concludes with a discussion of additional functionality and related software.

@@

# Description
- maybe some overview of spatial models, either here in intro. Contrast with GAMs, and highlight explicit estimation of flexible spatial covariance functions

@@

Though `sdmTMB` has been developed with spatial or spatiotemporal modelling as a primary focus, all of the statistical models it fits fall under the class of generalized linear mixed models (GLMMs).
GLMMs are available in several widely used packages, including `glmmTMB` (Brooks et al. 2017) and `lme4` (Bates et al. 2015), and to allow comparison we also allow non-spatial models to be constructed with `sdmTMB`.

We use the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time.
The general form of a GLMM is 
$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$. 
<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. 
The full `sdmTMB` model is
$$
g(\mu_{\bm{s},t}) = 
\bm{X}_{1,\bm{s},t} \bm{\beta} + 
\bm{\alpha}_g +
\bm{X}_{2,\bm{s},t} \gamma_t + 
\bm{X}_{3,\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} + 
\epsilon_{\bm{s},t},
$$
where

@@

The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a \mbox{Mat\'ern} covariance function. 
The \mbox{Mat\'ern} defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
(\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
a scaling parameter that is estimated (REF). 
The parameter $\nu$ controls the smoothness of the covariance function. 
We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].

By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in `sdmTMB` allow them to be modelled as a random walk or first-order autoregressive (AR1) process.
We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: `sdmTMB` allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in `mgcv` (Wood 2003), or to be modelled with threshold shapes (e.g., hockey stick models; Barrowman et al. 2003).

Highlights?

- Basic spatial model structure and notation / INLA / SPDE / GMRF defined. Advantages of PC-priors [], though sdmTMB also allows priors on most parameters (e.g. fixed effects, obs error variances) 
- Flexible extension of glmmTMB [Brooks et al. 2017] -- and includes non-spatial models for direct comparisons with glmmTMB [@osgood-zimmerman2021]
- Spatiotemporal models (RW/AR v IID)
- Time varying parameters
- Non-linear effects (smooth splines, threshold/breakpoint functions)
- Variance estimation and derived quantities (spatial range, area occupied etc)

>>>>

@@

@@

Here, we introduce our R package sdmTMB, which implements spatial and spatiotemporal GLMMs using TMB and INLA.
Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of spatial and spatiotemporal models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through X case studies, and concludes with a discussion of additional functionality and related software.

@@

# Description

Spatial and spatiotemporal models constructed with sdmTMB may be viewed as spatially explicit models, in the sense that they estimate parameters of a user-specified spatial covariance function, and these estimates may be converted into interpretable quantities common in spatial statistics (spatial decay, range). 
We contrast these approaches with semi- and non-parametric approaches that do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], some smooths in mgcv [@wood2017a]). By default, sdmTMB implements the Matérn covariance function (REF) (cite Rue et al?)---this function can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in a latitudinal direction may be different than in the longitudinal direction).

@@

Although sdmTMB has been developed with spatial or spatiotemporal modelling as a primary focus, the statistical models it fits fall under the class of GLMMs.
GLMMs are available in several widely used packages, including glmmTMB [@brooks2017], mgcv [@wood2017a], and lme4 [@bates2015], and to allow comparison with these packages, sdmTMB allows for spatial and spatiotemporal effects to be omitted.

By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in sdmTMB allow them to be modelled as a random walk or first-order autoregressive, AR(1), process.
We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: sdmTMB allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000B].

<!-- Paragraph on model selection here-->
Following glmmTMB, we allow AIC to be calculated for each sdmTMB model (Akaike XXXX). 
Because of well-documented biases of AIC with hierarchical or mixed-effects models (can cite Vaida and Blanchard or another), we also provide functionality to perform estimation via k-fold cross-validation. 
Our function `sdmTMB_cv()` allows for any number of folds or 'blocks', with assignments being user-specified or chosen randomly. 
The predictive log-density is calculated for each fold, and then summarized across folds for a total measure of the predictive accuracy of a model.

<!-- Paragraph on derived quantities here-->
The results from a model fit with sdmTMB can be used to generate quantities of interest, especially relevant to population status. 
Estimated parameters from the Matérn covariance function can be used to describe the spatial variance and Matérn range. 
We provide additional functionality for generating population-level summaries for each time slice; these metrics include the fraction of area occupied (TODO?), 2-dimensional center of gravity (`sdmTMB::get_cog()`), and total population index ( `sdmTMB::get_index()`).

<!--
Highlights?

- Basic spatial model structure and notation / INLA / SPDE / GMRF defined. Advantages of PC-priors [], though sdmTMB also allows priors on most parameters (e.g. fixed effects, obs error variances)
- Flexible extension of glmmTMB [Brooks et al. 2017] -- and includes non-spatial models for direct comparisons with glmmTMB [@osgood-zimmerman2021]
- Spatiotemporal models (RW/AR v IID)
- Time varying parameters
- Non-linear effects (smooth splines, threshold/breakpoint functions)
- Variance estimation and derived quantities (spatial range, area occupied etc)
-->

# Supporting Information

We use the notation of a bold capital letter to represent a matrix, a bold lowercase greek letter to represent a parameter vector, the index $\bm{s}$ to represent a point in space (x and y and coordinates), and the index $t$ to represent a point in time.
The general form of a GLMM is
$g(\bm{\mu}) = \bm{X \beta} + \bm{Z \psi}$,
with the inverse link function $g(x)$ relating the mean $\bm{\mu}$ to fixed-effect predictors in the design matrix $\bm{X}$ with estimated coefficients $\bm{\beta}$ and random effects with the design matrix $\bm{Z}$ with estimated coefficients $\bm{\psi}$.
<!-- The matrix $\bm{Z}$ allows random effects to be structured (e.g., intercepts varying by group) or unstructured, with $\bm{u}$ representing random variation. -->
For clarity, we partition the random effects $\bm{Z \psi}$ into a number of more interpretable components, each of which can optionally be included in the model. Though in practice it is rare to include all of these parameters, the full `sdmTMB` model is
$$
g(\mu_{\bm{s},t}) =
\bm{X}_{1,\bm{s},t} \bm{\beta} +
\bm{\alpha}_g +
\bm{X}_{2,\bm{s},t} \gamma_t +
\bm{X}_{3,t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t},
$$
where

@@

The random fields are constructed as Gaussian Markov random fields (GMRF), where the random effects are structured as MVN constrained by a \mbox{Mat\'ern} covariance function.
<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
The \mbox{Mat\'ern} defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
(\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
a scaling parameter that is estimated (REF).
The parameter $\nu$ controls the smoothness of the covariance function.
We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
